Stack Safety as a Security Property

### 2020-07-05 main.tex

The call stack is a perennially popular target for attacks
exploiting memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
hijacking of the program's control flow. To prevent such attacks, many
software and hardware techniques for stack safety have been proposed, % in the literature and implemented in production systems,
including stack canaries~\citep{Cowan+98},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
split stacks~\citep{Kuznetsov+14},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from nick;there could be more}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion of safety
is usually defined negatively, by enumerating the bad things that a safe system
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits 
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ... 
We are only aware of one, quite recent, attempt to frame a positive definition.
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, local state encapsulation and well-bracketed control flow.
They formalize these properties by defining an abstract capability-based machine   
that ``satisfies [stack safety] by construction.'' We believe there is still room
to craft a more abstract and less intensional characterization of
stack safety, and that is the goal of the present paper. \apt{still needs tuning}

Our approach is to give alternative formal
definitions of local state encapsulation and well-bracketed control flow
using the conceptual framework of language-based security.
%
The underlying notion behind local state encapsulation is the
intention to protect callers from callees: a callee should be not be
able to access or overwrite a caller's private data. We formalize this
intuition as a security property with both confidentiality and
integrity components.
%
Intuitively, {\em stack confidentiality} promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.  \bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}

\bcp{BCP idea: contrast intrinsic vs. extrinsic property definition
     Remember to talk about spatial vs temporal safety}

While integrity and confidentiality are standard security
properties, applying them in the stack-safety setting raises a novel challenge:
the notion of ``attacker'' changes during execution. Not only
does a caller need to be protected from its callee, but the callee
itself needs to be protected from {\em its} callees.
We address this challenge by adjusting the integrity and confidentiality
of different stack frames at each call or return transition between functions.\apt{connection to ``attacker'' still vague.}
\rb{Mention to varying ``accessing power'' of the attacker?
  The notion of attacker changes on certain actions that are not externally
  visible.}

@@

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  Abstract formal properties capturing the 
  intuitive concepts of local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in a very general setting. These properties
  rule out common examples of stack memory abuse.
\item
  A case study of the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:enforcement}).
  This policy enforces our stack safety property, and also a (provably) stronger property that is easier to test.
  \apt{If we have no testing, do we even want to talk about the testing property?}
\item
  A refinement of the case study that considers the more efficient Lazy Tagging and Clearing policy 
  of \citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:lazy}).  This policy does not enforce our property, but we give a
  variant of the policy that does enforce a (provably) weaker variant of our property. \apt{Awkward, but the truth I fear.}

@@

\section{Running Example}
\label{sec:running-example}

To help keep things concrete throughout the more technical sections
that follow, we first introduce a running example. We will
use it to demonstrate various potentially problematic
behaviours in this section, and then show how our properties and
enforcement mechanisms preclude them later on in the paper.

The program appears in \cref{fig:running-program}. It consists
of a {\tt main} function that performs a call to some function {\tt
  f}, which in turns makes a nested call to some function {\tt g}
before returning, all while performing some simple arithmetic
in the process.

\begin{figure}[b]
  \begin{centering}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 42, y = 0;
  y = g(0);
  print (x + y);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
g(z) = {
  int w = h(17);
  print z;
  return (z + w);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
h(v) = {
  return (v + 1);
}

\end{verbatim}
    \end{minipage}
\end{centering}
\caption{High-level Program with Simple Nested Calls}
\label{fig:running-program}
\end{figure}

@@

@@

    \row{\jal ~ \rra ~ 40}{??}{PC \leftarrow 40; ~ \rra \leftarrow \thepcctr}

@@

\begin{figure}
\[
\begin{array}{c|c|c|c|c}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 2}{[\rsp - 2] \leftarrow r_4}{[107] \leftarrow 18}
      {Store $r_4$ as the return value}
  \row{\lw ~ \rra ~ \rsp ~ \negate 1}{\rra \leftarrow [\rsp - 1]}{\rra \leftarrow 31}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 1}{\rsp \leftarrow \rsp - 1}{\rsp \leftarrow 108}
      {Restore the stack pointer of {\tt g}}
  \row{\jalr ~ \rra ~ \rra ~ 0}{??}{PC \leftarrow 31}
      {Return to {\tt g}}
  \end{array}
\]

>>>>

The call stack is a perennially popular target for attacks
exploiting memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, many
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from
  nick;there could be more}\bcp{How about static techniques??  We decided
  these were also in-scope.}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion of safety
is usually defined negatively, by enumerating the bad things that a safe system
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
We are only aware of one, quite recent, attempt to frame a positive definition.
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.'' We believe there is still room
to craft a more abstract and less intensional characterization of
stack safety, and that is the goal of the present paper. \apt{still needs
  tuning}\bcp{Yes, I think we need to go into a little more detail on what
  their definition looks like in order to justify why an alternative
  definition might be valuable.  What questions does their presentation not
  answer? What uses does it not support? E.g., can it be enforced
  efficiently? (I guess so, by capabilities, though I think their
  capabilities are more powerful than what Cheri actually provides, while
  our policies are exactly what Nick needs.)  Why would it be good to look
  at stack safety through the lens of security?  (Because it is
  fundamentally a security mechanism?)}
\rb{BCP idea: contrast intrinsic vs. extrinsic property definition.
     Remember to talk about spatial vs temporal safety}

Our main contribution in this paper is an alternative formal presentation of
local state encapsulation and well-bracketed control flow using the
conceptual and technical framework of language-based security.
%
The underlying notion behind local state encapsulation is the intention to
protect callers from callees: a callee should be not be able to access or
overwrite private data in the caller's stack frame. We formalize this
intuition as a security property with both confidentiality and integrity
components.
%
Intuitively, {\em stack confidentiality} promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.  \bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}

While integrity and confidentiality are standard security
properties, applying them in the stack-safety setting raises a novel challenge:
the notion of ``attacker'' changes during execution. Not only
does a caller need to be protected from its callee, but the callee
itself needs to be protected from {\em its} callees.
We address this challenge by adjusting the integrity and confidentiality
of different stack frames at each call or return transition between functions.\apt{connection to ``attacker'' still vague.}
\rb{Mention to varying ``accessing power'' of the attacker?
  The notion of attacker changes on certain actions that are not externally
  visible.} \bcp{Maybe the best way is to grab the bull by the horns: ``If
  we're going to phrase stack safety as a security property, we need to talk
  about an attacker model, and this is where the technicalities get really
  interesting...''}

@@

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  Abstract formal properties capturing the
  intuitive concepts of local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in a very general setting. These properties
  rule out common examples of stack memory abuse.
\item
  A case study of the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:enforcement}).
  This policy enforces our stack safety property, and also a (provably) stronger property that is easier to test.
  \apt{If we have no testing, do we even want to talk about the testing
    property?} \bcp{Not here; just in future work.  I guess we need to see
    whether we wind up with any results at all about testing.}
\item
  A refinement of the case study that considers the more efficient Lazy Tagging and Clearing policy
  of \citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:lazy}).  This policy does not enforce our property, but we give a
  variant of the policy that does enforce a (provably) weaker variant of our
  property.
%
  \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
    but the point we can make is still strong: To make the property easier
    to enforce, we can slightly weaken its guarantees; the weaker version is
    a bit more complicated to state but still retains the pragmatic
    protections of the original.}

@@

\section{Stack Safety and Unsafety by Example}
\label{sec:running-example}

We begin with a running example that will serve both to concretely
demonstrate the ``stack-unsafe'' behaviours we intend to preclude and to
ground and illustrate the formal properties and enforcement mechanisms in
later sections.

\subsection{Running Example}

The example program appears in \Cref{fig:running-program}. It consists of a
{\tt main} function, written in some C-like high-level language, that
performs a call to some function {\tt f}, which in turns makes a nested call
to some function {\tt g} before returning, performing some simple arithmetic
in the process.

\begin{figure}[b]
\hfill
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 42, y = 0;
  y = g(0);
  print (x + y);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
g(z) = {
  int w = h(17);
  print z;
  return (z + w);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
h(v) = {
  return (v + 1);
}

\end{verbatim}
    \end{minipage}
\caption{A High-level Program With Nested Calls}
\label{fig:running-program}
\end{figure}

@@

\bcp{A very small nit: Change
  $\rsp-2$ to $\rsp\mathord{-}2$}

@@

    \row{\jal ~ \rra ~ 40}{??}{\PCname \leftarrow 40; ~ \rra \leftarrow \thepcctr}

@@

\begin{figure}
\[
\begin{array}{c|c|c|c|c}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 2}{[\rsp - 2] \leftarrow r_4}{[107] \leftarrow 18}
      {Store $r_4$ as the return value}
  \row{\lw ~ \rra ~ \rsp ~ \negate 1}{\rra \leftarrow [\rsp - 1]}{\rra \leftarrow 31}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 1}{\rsp \leftarrow \rsp - 1}{\rsp \leftarrow 108}
      {Restore the stack pointer of {\tt g}}
  \row{\jalr ~ \rra ~ \rra ~ 0}{\leo{??}}{\PCname \leftarrow 31}
      {Return to {\tt g}}
  \end{array}
\]
